---
title: "Problem Set #1: Exploring Wage Differences in the U.S. Early Care and Education Workforce"
output: html_document
date: "2026-08-03"

##
## Problem Set 1 
## Data Prep & Descriptives: NSECE 2019 Workforce Questionaire
## Katia Bouali
## R Version 4.5.2
##


---
```{r Set up}
## Set up 
library(dplyr)
library(tidyr)
library(tidyverse)
library(forcats)
library(ggplot2)


describe_num <- function(x) {
  x <- x[!is.na(x)]
  data.frame(
    n = length(x), mean = mean(x), sd = sd(x), median = median(x),
    min = min(x), max = max(x), range = max(x) - min(x),
    skew = mean((x - mean(x))^3) / sd(x)^3
  )
}

freq_table <- function(x) {
  x <- x[!is.na(x)]
  tab <- table(x)
  data.frame(category = names(tab), n = as.integer(tab),
             pct = round(100 * as.integer(tab) / length(x), 1))
}


## Load data
load("~/UChicago_Student/2_Data_For_Policy_Analys_-_Mgmt/Data/37941-0005-Data.rda")
nsece_raw <- da37941.0005 

## Subset to variables
nsece <- nsece_raw %>%
  select(
    WF9_WORK_WAGE,
    WF9_WORK_ROLE,
    WF9_WORK_YRS,
    WF9_CHAR_HISP,
    WF9_CHAR_RACE,
    WF9_CHAR_EDUC,
    WF9_CHAR_GENDER
  )

## Rename variables
nsece <- nsece %>%
  rename(
    wage       = WF9_WORK_WAGE,
    role       = WF9_WORK_ROLE,
    yrs_worked = WF9_WORK_YRS,
    hispanic   = WF9_CHAR_HISP,
    race       = WF9_CHAR_RACE,
    educ       = WF9_CHAR_EDUC,
    gender     = WF9_CHAR_GENDER
  )

## Missing data

nsece %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  mutate(pct_missing = round(100 * n_missing / nrow(nsece), 1))


## Recode
# role, hispanic, race,gender: nominal -> leace as unordered factors (already are)
# edu, yrs_worked: ordinal -> set explixit order so later tests treat them correctly


nsece <- nsece %>%
  mutate(
    educ = factor(educ, levels = c(
      "(1) Less than High School",
      "(3) GED or high school equivalency",
      "(4) High school graduate",
      "(5) Some college credit but no degree",
      "(6) Associate degree (AA, AS)",
      "(7) Bachelor's degree (BA, BS, AB)",
      "(8) Graduate or professional degree"
    ), ordered = TRUE),
    
    yrs_worked = factor(yrs_worked, levels = c(
      "(01) Less than half a year (0-6 months)", "(02) 7-12 months",
      "(03) 1 year", "(04) 2 years", "(05) 3 years", "(06) 4 years",
      "(07) 5 years", "(08) 6 years", "(09) 7 years", "(10) 8 years",
      "(11) 9 years", "(12) 10 years", "(13) 11 to 15 years",
      "(14) 16 to 20 years", "(15) 21 to 25 years", "(16) 26 or more years"
    ), ordered = TRUE)
  )

## Level of measurment/variable type

# wage        -> interval/ratio, continuous   (hourly dollar amount; true zero)
# role        -> nominal, discrete            (3 categories: Aide/asst, Teacher/lead, Other)
# yrs_worked  -> ORDINAL, discrete            (16 bins of unequal width, e.g. "11 to 15 years" -
#                                               NOT a raw count, so treat as ordinal, not ratio)
# hispanic    -> nominal, discrete            (binary)
# race        -> nominal, discrete            (4 categories: White, Black, Asian, Other)
# educ        -> ORDINAL, discrete            (7 ranked categories, unequal spacing)
# gender      -> nominal, discrete            (binary)

## Descriptive Statistics

```
**Question 1: Data Preparation and Descriptive Statistics**

*The analytic sample was drawn from the NSECE 2019 Questionnaire (N=5,192) and subset to seven variables: hourly wage, staff role, years worked in the program, Hispanic/Latino ethnicity, race, educational attainment, and gender. Each variable was classified by level of measurement to guide the choice of descriptive statistics and, layer, inferential tests.*

**Hourly wage** *is a continous interval/ratio variable. Among the 4,270 respondents with vald wage data (17.8% missing), the mean hourly wage was \$15.33 (SD = \$10.65), while the median was noticeably lower at \$12.83, with a range from \$1.29 to \$107.50. As shown in Figure 1, the distribution is strongly right-skewed (skewness \~ 5), with a long tail of high earners pulling the mean above the median. The median is therefore the more representiative measure of central tendency for this variable, and it will also be preferred in wage-difference tests where the distributional assumptions of mean-based tests may not hold.*

```{r Figure 1}
# -- Wage (continuous) -- Figure 1. Distribution of hourly wage.


ggplot(nsece, aes(x = wage)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white", na.rm = TRUE) +
  labs(title = "Distribution of Hourly Wage", x = "Hourly Wage ($)", y = "Count") +
  theme_minimal()
```

**Staff role** *is a nominal, discrete variable with three categories. As Figure 2 shows, the workforce sample was composed primarily of teachers/lead teachers (65.4%) and aides/assistant teachers (34.3%), with an "other/undetermined" category accounting for a negligible 0.4% (n=18)*

```{r}

# -- Role (nominal) -- Figure 2. Distribution of staff role.
freq_table(nsece$role)

ggplot(nsece %>% filter(!is.na(role)), aes(x = fct_infreq(role))) +
  geom_bar(fill = "coral") +
  labs(title = "Staff Role", x = NULL, y = "Count") +
  coord_flip() +
  theme_minimal()
```

**Years worked in the program** *was recorded in 16 ordinal, unequal-width bins rather than ratio-level. Figure 3 illustrate a distribution right-skewed toward shorter tenures: roughly a third of respondants (33.7%) reported less than two years in their current program, while only about 10% reported more than 15 years.*

```{r}

# -- Years worked (ordinal) -- Figure 3. Distribution of years worked in program.
freq_table(nsece$yrs_worked)

ggplot(nsece %>% filter(!is.na(yrs_worked)), aes(x = yrs_worked)) +
  geom_bar(fill = "darkseagreen") +
  labs(title = "Years Worked in Program", x = NULL, y = "Count") +
  coord_flip() +
  theme_minimal()
```

**Hispanic/Latino ehtnicity** *and* **gender** *are both nominal, discrete, binary variables. Approximately 24.4% of respondents identified as Hispanic or Latino. The sample is heavily skewed by gender, with 97.6% identifying as female and only 2.4% male (n = 112)- a notable imbalance to keep in mind for any later gender comparison, since the male subgroup is small.*

```{r}

# -- Hispanic/Latino (nominal) -- 
freq_table(nsece$hispanic)


# -- Gender (nominal) --
freq_table(nsece$gender)
```

**Race** *is a nominal, dsicrete variable with four categories. (18.3% missing). As shown in figure 4, the sample was predominantly White (66.6%), followed by Black or African American (23.4%), Other (5.1%), and Asian (5.0%).* *Note that the race variable is derived from a survey item allowing respondents to select multiple races, and respondents selecting more than one race were coded into the 'other' category, so racial comparisons in this analysis reflect single-race groupings versus a herterogenous 'other' category, not a full accounting of racial identification.*

```{r}
# -- Race (nominal) -- Figure 4. Distribution of race.
freq_table(nsece$race)

ggplot(nsece %>% filter(!is.na(race)), aes(x = fct_infreq(race))) +
  geom_bar(fill = "goldenrod") +
  labs(title = "Race", x = NULL, y = "Count") +
  coord_flip() +
  theme_minimal()
```

**Educational attainment** *is an ordinal, discrete variable with seven ranked categories. Figure 5 shows a distribution concentrated in the middle categories, with 27.3% reporting some college credit but no degree, 24.9% holding a BA degree, and 17.8% holding an associate degree. Very few respondents (1.8%) reported less than a high school education.*

```{r}
# -- Education (ordinal) -- Figure 5. Distribution of educational attainment.
freq_table(nsece$educ)

ggplot(nsece %>% filter(!is.na(educ)), aes(x = educ)) +
  geom_bar(fill = "plum") +
  labs(title = "Educational Attainment", x = NULL, y = "Count") +
  coord_flip() +
  theme_minimal()
```

*Missing data ranged from about 9% (role, education) to 18% (wage,race). A cross-check of wage missingness by staff role found no evidence of concentration in a particular group (missingness ranged narrowly from 8.8% to 11.1% across roles), suggesting the missingness is unlikely to introduce substantial bias into the wage analyses that follow*

**Figure 1. Distribution of hourly wage. Figure 2. Distribution of staff role. Figure 3. Distribution of years worked in program. Figure 4. Distribution of race. Figure 5. Distribution of educational attainment.**

**Question 2: Significance Tests- Hourly Wage and Workforce Characteristics** *Because wage is strongly right-skewed (skewed \~ 5 --\> mean \$15.33 vs. median \$12.83), it violates the normality assumption behind t-tests, ANOVA, and Pearson correlation. Nonparametric tests, which compare ranks or distributions rather than means, were used consistently across all give analyses below.*

**Wage and Staff Role** *A Kruskal-Wallis test found a statistically significant difference in wage across staff roles, x\^(2)=107.75, p\<.001. Median hourly wage was lowest for aides/assistant teachers (\$12.00, m=1,469), higher for teachers/lead teachers (\$13.40, n=2,785), and highest for the small "other/undetermined" group (\$15.50, n=16). This indicates that staff role is significantly associated with pay, with support level positions earning less than lead teaching positions.*

```{r}

kruskal.test(wage ~ role, data = nsece)

nsece %>%
  filter(!is.na(role), !is.na(wage)) %>%
  group_by(role) %>%
  summarise(median_wage = median(wage), n = n()) 

```

**Wage and Years worked** *A Spearman rank correlation showed a statistically significant, positive but modest association between wage and years worked in the program, (rho)=.28, p\<.001. More tenure is associated with higher pay, bit the relationship is weak to moderate, suggesting experience alone does not strongly predict wage growth in this field.*

```{r}

nsece_b <- nsece %>% filter(!is.na(wage), !is.na(yrs_worked))

cor.test(nsece_b$wage, as.integer(nsece_b$yrs_worked), method = "spearman")
cor.test(nsece_b$wage, as.integer(nsece_b$yrs_worked), method = "spearman", exact = FALSE)
```

**Wage and Race** *A Kruskal-Wallis test found a statistically significant difference in wage by race, x\^2(3)=131.11, p\<.001. Median wage was lowest for Black educators (\$11.50, n=882), followed by white (\$13.00, n=2,607) and other race educators (\$13.00, n=204), with asian educators earning the highest median wage at (\$16.00, n=186). This points to a meaningful racial wage gap within the ECE workforce.*

```{r}

kruskal.test(wage ~ race, data = nsece)

nsece %>%
  filter(!is.na(race), !is.na(wage)) %>%
  group_by(race) %>%
  summarise(median_wage = median(wage), n = n())
```

**Wage and Educational Attainment** *A Spearman rank correlation showed a statistically significant, oderate positive association between wage and educational attainment, (rho)=.48, p\<.001 --\> which shows to be the strongest relationship among the ordinal predictors tested. Higher educational attainment is associated with meaningfully higher hourly wages.*

```{r}

nsece_d <- nsece %>% filter(!is.na(wage), !is.na(educ))
cor.test(nsece_d$wage, as.integer(nsece_d$educ), method = "spearman")
cor.test(nsece_d$wage, as.integer(nsece_d$educ), method = "spearman", exact = FALSE)
```

**Wage and Gender** *A Wilcoxon rank-sum (Mann-Whitney) test found a statistically significant differnce in wage by gender, w=222,296, p=.043. Median wage was somewhat higher for men (\$13.50, n=96) than woman (\$12.80, n=4,133). However, given the extreme imbalance in group sizes (the sample is 97.6% female), this result should be cautiously interpreted rather than treated as strong evidence of a robust gender wage gap.*

```{r}

wilcox.test(wage ~ gender, data = nsece)

nsece %>%
  filter(!is.na(gender), !is.na(wage)) %>%
  group_by(gender) %>%
  summarise(median_wage = median(wage), n = n())
```

**Methods** *Choice of statistical methods--\> Two families of significance tests are commonly used ofr the relationships examined in Q2: parametric tests (Pearson correlation, independent samples t-test, one-way ANOVA) and their nonparametric counterparts Spearman rank correlation, Wilcoxon Mann Whitney test, Kruskal Wallis test). The parametric tests assume the outcome variable is approximately normally distributed within each group or across the relationship being tested. violations of this assumption can distort p-values and standard errors, particularly when a distribution is heavily skewed. As established in Q1, hourly wage is strongly right skewed (skewness \~ 5, mean \$15.33 vs median \$12.83) which violates the normality assumption underlying the parametric tests.*

*For this reason, nonparametric alternatives were used throughout Q2, consistent with the guidance in Agresti (2018): nonparametric methods "are also useful when the two groups have highly skewed distributions, because then the mean may not be a meaningful summary measure" (Ch 7.7) Specifically:*

*The Kriskal Wallis test was used in place of one-way ANOVA for comparing wage across staff role (3 groups) and race (4 groups). Rather than comparing group means (as ANOVA does), Kruskal Wallis ranks all observations and compares mean ranks across groups, making it robust to the skew and outliers present in the wage variable (Agresti, 2018 Ch.12.3)*

*The Spearman rank correlation was used in place of Pearson correlation for the relationsips between wage and the two ordinal variables (years worked, educational attainment). Pearson correlation assumes a linear relationship between two continous, normally distributed variables. Spearman instead measures the strength of a monotonic relationship using ranks, which is appropriate both because wage is skewed and because years worked and education are ordinal rather than truly continuous variables*

*The Wilcoxon rank sum (Mann Whitney) test was used in place of an independent-samples t-test for comparing wage between genders. Like Kriskal Wallis, it compares rank distributions rather than means, and is the standard nonparametric substitue for a two-group test (Agresti, 2018, Ch 7.7)*

*Using nonparametric tests consistently across all five analyses also has a secondary benefit--\> it keeps the choice of test uniform across Q2 rather than switching justification variable by variable and it also aligns with reporting medians, rather than means, as the primary descriptive statistic for wage establishment in Q1.*

**Question 3: Policy Implications** *Taken togeher, the analyses point to several segments of the ECE workforce most affected by low wages. First, staff role matters: aides and assitant teachers earn a significantly lower median wage than teachers and lead teachers (p\<.001), suggesting compensation policy should not treat "ECE educator" as a single undifferentiated category --\> entry-level and support staff face a distinct wage disadvantage. Second, race is strongly associated with wage: Black educators earn a notably lower median wage than their white, asian, and other race counterparts (p\<.001), indicating that racial age disparities are a persistent and significant feature of this workforce, not incidental variation. Third, education attainment shows the strongest positive association with wage of the ordinal predictors tested (p=.48, p\<.001), meaning workers with lower levels of formal education are systematically concentrated among the lowest earners, even though the ECE workforce already has substantial postsecondary attainment relative to its pay. By contrast, years of experience shows only a weak positive relationship with wage (p=.28), suggesting that tenure alone does little to lift wages, a structural problem, since it implies the field offers limited return on staying in the profession long term. Evidence for a gender wage gap was comparatively weak: while the Man-Whitney test reached conventional significance (p=.043), the male subgroup was very small (n=96 vs 4,133 women), so this result should be interpreted catiously rather than treated as a robust finding.*

*Given these findings and patterns, policy interventions aimed at improving ECE compensation should prioritize: 1. raising wage floors specifically for aide and assistant-teacher roles rather than only lead teachers, sine role-based stratification is one of the clearest drivers of low pay. 2. directly addressing the racial wage gap affecting Black educators, since this disparity persists independent of role or education, and 3. building compensation structures that reward educational credentials more consistently and reward tenure more meaningfully, since the current wage structure does relatively little to compensate experience. Given the very small male sub sample, any policy response to a possible gender gap would benefit from a larger, more balanced sample before being prioritized.* *It is worth noting that the 'other' race category combines several distinct groups (including multiracial respondents) so this racial wage gap analysis understates the full complexity of racial disparities in the ECE workforce and should be treated as a starting point rather than a complete picture*
